---
title: "rhandsontable Introduction"
author: "Jonathan Owen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rhandsontable Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r echo = FALSE}
library(rhandsontable)
library(knitr)

opts_knit$set(warning = FALSE, error = FALSE, message = FALSE, cache = FALSE,
              fig.width=7, fig.height=3)
```

## Introduction

  rhandsontable is a htmlwidget based on the [handsontable.js](https://www.handsontable.com) library.

> Handsontable is a data grid component with an Excel-like appearance. Built in JavaScript, it integrates with any data source with peak efficiency. It comes with powerful features like data validation, sorting, grouping, data binding, formula support or column ordering.
([via](https://www.handsontable.com))

## Column Types

The table includes support for numeric, logical, character and Date types.  Logical values will appear as check boxes, and the [pikaday.js](https://github.com/dbushell/Pikaday) library is used to specify Date values.

rhandsontable attempts to map R classes to an appropriate handsontable type.  Factors will be mapped to `dropdown`, with the choices specified by `level` and `allowInvalid` set to `FALSE`.  To allow new levels, set `allowInvalid` to `TRUE` (using `hot_col`; it may also be desirable to set `strict` to `FALSE`).  When running in `shiny`, using `hot_to_r` will preserve custom factor ordering, and if new levels are allowed, they will be added to the end.

```{r}
DF = data.frame(integer = 1:10,
                   numeric = rnorm(10),
                   logical = rep(TRUE, 10), 
                   character = LETTERS[1:10],
                   factor = factor(letters[1:10], levels = letters[10:1], 
                                   ordered = TRUE),
                   factor_allow = factor(letters[1:10], levels = letters[10:1], 
                                         ordered = TRUE),
                   date = seq(from = Sys.Date(), by = "days", length.out = 10),
                   stringsAsFactors = FALSE)

rhandsontable(DF, width = 600, height = 300) %>%
  hot_col("factor_allow", allowInvalid = TRUE)
```

To improve readability, `NA` values will be displayed as blank cells.  This requires converting columns containing `NA` to characters, and in the case of factors and Dates, may not display the data in the desired format.  It may be beneficial to concert these type of columns to character before passing to `rhandsontable`.

```{r}
DF_na = data.frame(integer = c(NA, 2:10), 
                   logical = c(NA, rep(TRUE, 9)), 
                   character = c(NA, LETTERS[1:9]),
                   factor = c(NA, factor(letters[1:9])),
                   date = c(NA, seq(from = Sys.Date(), by = "days", 
                                    length.out = 9)),
                   stringsAsFactors = FALSE)

DF_na$factor_ch = as.character(DF_na$factor)
DF_na$date_ch = c(NA, as.character(seq(from = Sys.Date(), by = "days", 
                                       length.out = 9)))

rhandsontable(DF_na, width = 550, height = 300)
```

### Dropdown / Autocomplete

To control character column values, the column type can be specified as `dropdown` or `autocomplete`.

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

# try updating big to a value not in the dropdown
rhandsontable(DF, rowHeaders = NULL, width = 550, height = 300) %>%
  hot_col(col = "big", type = "dropdown", source = LETTERS) %>%
  hot_col(col = "small", type = "autocomplete", source = letters,
          strict = FALSE)
```

### Password

A column can also be specified as a `password` type.

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, width = 550, height = 300) %>%
  hot_col("small", "password")
```

### Sparkline

New in version 0.2, [sparkline.js](http://omnipotent.net/jquery.sparkline/) charts can be added to the table.  Thanks to the sparkline package and Ramnath Vaidyanathan for inspiration.

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

DF$chart = c(sapply(1:5,
                    function(x) jsonlite::toJSON(list(values=rnorm(10),
                                                      options = list(type = "bar")))),
             sapply(1:5,
                    function(x) jsonlite::toJSON(list(values=rnorm(10),
                                                      options = list(type = "line")))))

rhandsontable(DF, rowHeaders = NULL, width = 550, height = 300) %>%
  hot_col("chart", renderer = htmlwidgets::JS("renderSparkline"))
```

### Custom Renderer

It's also possible to define a custom column renderer function.  For example, it may be desirable to include html in a cell.  The example below mimics [Custom renderers](http://docs.handsontable.com/0.16.1/demo-custom-renderers.html).

```{r fig.height = 5, fig.width = 8}
DF = data.frame(
  title = c(
    "<a href='http://www.amazon.com/Professional-JavaScript-Developers-Nicholas-Zakas/dp/1118026691'>Professional JavaScript for Web Developers</a>",
    "<a href='http://shop.oreilly.com/product/9780596517748.do'>JavaScript: The Good Parts</a>",
    "<a href='http://shop.oreilly.com/product/9780596805531.do'>JavaScript: The Definitive Guide</a>"
  ),
  desc = c(
    "This <a href='http://bit.ly/sM1bDf'>book</a> provides a developer-level introduction along with more advanced and useful features of <b>JavaScript</b>.",
    "This book provides a developer-level introduction along with <b>more advanced</b> and useful features of JavaScript.",
    "<em>JavaScript: The Definitive Guide</em> provides a thorough description of the core <b>JavaScript</b> language and both the legacy and standard DOMs implemented in web browsers."
  ),
  comments = c(
    "I would rate it &#x2605;&#x2605;&#x2605;&#x2605;&#x2606;",
    "This is the book about JavaScript",
    "I've never actually read it, but the <a href='http://shop.oreilly.com/product/9780596805531.do'>comments</a> are highly <strong>positive</strong>."
  ), 
  cover = c(
    "http://ecx.images-amazon.com/images/I/51bRhyVTVGL._SL50_.jpg",
    "http://ecx.images-amazon.com/images/I/51gdVAEfPUL._SL50_.jpg",
    "http://ecx.images-amazon.com/images/I/51VFNL4T7kL._SL50_.jpg"
 ),
 stringsAsFactors = FALSE
)

rhandsontable(DF, allowedTags = "<em><b><strong><a><big>", 
              width = 800, height = 450, rowHeaders = FALSE) %>%
  hot_cols(colWidths = c(200, 200, 200, 80)) %>%
  hot_col(1:2, renderer = "html") %>%
  hot_col(1:3, renderer = htmlwidgets::JS("safeHtmlRenderer")) %>%
  hot_col(4, renderer = "
    function(instance, td, row, col, prop, value, cellProperties) {
      var escaped = Handsontable.helper.stringify(value),
        img;
  
      if (escaped.indexOf('http') === 0) {
        img = document.createElement('IMG');
        img.src = value;
  
        Handsontable.Dom.addEvent(img, 'mousedown', function (e){
          e.preventDefault(); // prevent selection quirk
        });
  
        Handsontable.Dom.empty(td);
        td.appendChild(img);
      }
      else {
        // render as text
        Handsontable.renderers.TextRenderer.apply(this, arguments);
      }
  
      return td;
    }")
```

For `shiny` apps, use `renderer = htmlwidgets::JS("safeHtmlRenderer")` to display columns with html data.  The allowed html tags default to `<em><b><strong><a><big>`, but the (hidden) `allowedTags` parameter can in `rhandsontable` can be used to customize this list.

## Right-Click Menu

Right-clicking in a cell will enable a context menu that includes customizable table actions via the `hot_context_menu` function.  For shiny apps, formatting and comment updates made via the context menu are not currently retained.

To disable the context menu, set `contextMenu = FALSE` in `hot_table` (or `rhandsontable`).

### Add / Remove Rows & Columns

By default a user can add or remove table rows and columns, but this functionality can be disabled.  Note that Handsontable does not allow column be added or deleted to the table if column types are defined (i.e. `useTypes == TRUE` in `rhandsontable`).

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, width = 550, height = 300) %>%
  hot_context_menu(allowRowEdit = FALSE, allowColEdit = FALSE)
```

### Customizing

The `customOpts` parameter of `hot_context_menu` can be used to add custom functionality to the context menu.  Below are a couple examples.

#### Export to CSV

This example illustrates how to add an option to export the table to a csv file.

```{r}
MAT = matrix(rnorm(50), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:5]))

rhandsontable(MAT, width = 550, height = 300) %>%
  hot_context_menu(
    customOpts = list(
      csv = list(name = "Download to CSV",
                    callback = htmlwidgets::JS(
                      "function (key, options) {
                         var csv = csvString(instance);

                         var link = document.createElement('a');
                         link.setAttribute('href', 'data:text/plain;charset=utf-8,' +
                           encodeURIComponent(csv));
                         link.setAttribute('download', filename);

                         document.body.appendChild(link);
                         link.click();
                         document.body.removeChild(link);
                       }"))))
```

#### Search

This example illustrates how to enable the search functionality in Handsontable.

```{r}
DF = data.frame(val = 1:10,
                bool = TRUE,
                big = LETTERS[1:10],
                small = factor(letters[1:10]),
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, search = TRUE, width = 550, height = 300) %>%
  hot_context_menu(
    customOpts = list(
      search = list(name = "Search",
                    callback = htmlwidgets::JS(
                      "function (key, options) {
                         var srch = prompt('Search criteria');

                         this.search.query(srch);
                         this.render();
                       }"))))
```

Future enhancements will look to expand export options.

## Numeric Formatting

Numeric columns are formatted using the [numeral.js](http://numeraljs.com/) library.

```{r}
DF = data.frame(int = 1:10, float = rnorm(10), cur = rnorm(10) * 1E5,
                lrg = rnorm(10) * 1E8, pct = rnorm(10))

rhandsontable(DF, width = 550, height = 300) %>%
  hot_col("float", format = "0.0") %>%
  hot_col("cur", format = "$0,0.00") %>%
  hot_col("lrg", format = "0a") %>%
  hot_col("pct", format = "0%")
```

### Specify Locale

The `language` parameter for `hot_col` can be used to change the locale.  See the [numeral.js](http://numeraljs.com/) library for language options.

```{r}
DF = data.frame(dollar = rnorm(10), euro = rnorm(10), yen = rnorm(10))

rhandsontable(DF * 1000, width = 550, height = 300) %>%
  hot_col("dollar", format = "$0,000.00", language = "en") %>%
  hot_col("euro", format = "0,000.00 $", language = "de") %>%
  hot_col("yen", format = "$0,000.00", language = "ja")
```

## Read Only

The whole table and individual columns can to set to `readOnly` to prevent the user from making changes.

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, readOnly = TRUE, width = 550, height = 300) %>%
  hot_col("val", readOnly = FALSE)
```

## Sorting

Column sorting can be enabled; sorting only impacts the widget and will not reorder the original data set.

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, width = 550, height = 300) %>%
  hot_cols(columnSorting = TRUE)
```

## Highlight Rows & Columns

With larger tables it my be desirable to highlight the row and column for a selected cell.  

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

# click on a cell to see the highlighting
rhandsontable(DF, width = 550, height = 300) %>%
  hot_table(highlightCol = TRUE, highlightRow = TRUE)
```

## Sizing

Column and row dimensions can be customized.  For larger data sets, (multiple) top rows and left columns can be frozen.

```{r fig.height = 6, fig.width = 6}
MAT = matrix(rnorm(50), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:5]))

rhandsontable(MAT, width = 600, height = 600) %>%
  hot_cols(colWidths = 100) %>%
  hot_rows(rowHeights = 50)
```

### Row Header Width

The width of the row header column can be customized using `rowHeaderWidth`.

```{r fig.height = 6, fig.width = 6}
rhandsontable(mtcars, rowHeaderWidth = 200)
```

### Streching

The table can be streched to the full width by using `stretchH`.

```{r fig.height = 6, fig.width = 6}
MAT = matrix(rnorm(30), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:3]))

rhandsontable(MAT, width = 600, height = 300, stretchH = "all")
```

## Fixed Rows / Columns

For larger data sets, (multiple) top rows and left columns can be frozen.

```{r}
MAT = matrix(rnorm(26 * 26), nrow = 26, dimnames = list(LETTERS, letters))

# scroll through the table to see the fixed row and column
rhandsontable(MAT, width = 550, height = 300) %>%
  hot_cols(fixedColumnsLeft = 1) %>%
  hot_rows(fixedRowsTop = 1)
```

## Cell Comments

Comments (hover) can also be added to individual cells and will appear as red flags in the upper right of the cell. 

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, width = 550, height = 300) %>%
  hot_cell(1, 1, "Test comment")
```

Additionally, comments can be added via `data.frame` or `matrix`.

```{r}
MAT_comments = matrix(ncol = ncol(DF), nrow = nrow(DF))
MAT_comments[1, 1] = "Test comment"
MAT_comments[2, 2] = "Another test comment"

rhandsontable(DF, comments = MAT_comments, width = 550, height = 300)
```

Finally, comments can also be added via the right-click context menu, but these updates will not currently be retained by shiny.

## Borders

Custom borders can be drawn around cells to highlight specific items.  Borders can also be added via the right-click context menu, but these updates will not currently be retained by shiny.

```{r}
MAT = matrix(rnorm(50), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:5]))

rhandsontable(MAT, width = 550, height = 300) %>%
  hot_table(customBorders = list(list(
    range = list(from = list(row = 1, col = 1),
                 to = list(row = 2, col = 2)),
    top = list(width = 2, color = "red"),
    left = list(width = 2, color = "red"),
    bottom = list(width = 2, color = "red"),
    right = list(width = 2, color = "red"))))
```

## Validation

### Numeric Columns

Pre-defined validation can be added for numeric columns in two ways:

* specify a min and max and any values within the range to exclude
* similar to a `dropdown` column, specify allowed values

```{r}
MAT = matrix(rnorm(50), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:5]))

rhandsontable(MAT * 10, width = 550, height = 300) %>%
  hot_validate_numeric(col = 1, min = -50, max = 50, exclude = 40)

rhandsontable(MAT * 10, width = 550, height = 300) %>%
  hot_validate_numeric(col = 1, choices = c(10, 20, 40))
```

### Character Columns

For character columns, a vector of allowed options can be specified.  A more user-friendly approach may be to use a `dropdown` column with `strict = TRUE`.

```{r}
DF = data.frame(val = 1:10, bool = TRUE, big = LETTERS[1:10],
                small = letters[1:10],
                dt = seq(from = Sys.Date(), by = "days", length.out = 10),
                stringsAsFactors = FALSE)

rhandsontable(DF, width = 550, height = 300) %>%
  hot_validate_character(col = "big", choices = LETTERS[1:10])
```

### Custom

It is also possible to create a custom validation function in JavaScript.

```{r}
MAT = matrix(rnorm(50), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:5]))

# try to update any cell to 0
rhandsontable(MAT * 10, width = 550, height = 300) %>%
  hot_cols(validator = "
           function (value, callback) {
            setTimeout(function(){
              callback(value != 0);
            }, 1000)
           }",
           allowInvalid = FALSE)
```

## Conditional Formatting

Conditional formatting can also be specified via custom JavaScript function.  Future enhancements will look to simplify this interface.

```{r, fig.width = 8}
MAT = matrix(runif(100, -1, 1), nrow = 10,
             dimnames = list(LETTERS[1:10], LETTERS[1:10]))
diag(MAT) = 1
MAT[upper.tri(MAT)] = MAT[lower.tri(MAT)]
rhandsontable(MAT, readOnly = TRUE, width = 750, height = 300) %>%
  hot_cols(renderer = "
           function (instance, td, row, col, prop, value, cellProperties) {
             Handsontable.renderers.TextRenderer.apply(this, arguments);
             if (row == col) {
              td.style.background = 'lightgrey';
             } else if (col > row) {
              td.style.background = 'grey';
              td.style.color = 'grey';
             } else if (value < -0.75) {
              td.style.background = 'pink';
             } else if (value > 0.75) {
              td.style.background = 'lightgreen';
             }
           }")
```

## Heatmap

The [chroma.js](http://old.driven-by-data.net/about/chromajs/) library can be used to turn the table into a heatmap.

```{r}
MAT = matrix(rnorm(50), nrow = 10, dimnames = list(LETTERS[1:10],
                                                   letters[1:5]))

rhandsontable(MAT, width = 550, height = 300) %>%
  hot_heatmap()
```

## Shiny

The data grid will be editable by default and can be used as input to a `shiny` app.  A few `shiny` and `shinydashboard` example links are listed below.  Note that the shinyapps.io links may not work if the has hit the monthly usage limit.

When using `rhandsontable` and an input, these apps generally use the `hot_to_r` function to convert data to an R data.frame and `shiny::reactiveValues` to store a cached version of the data (see the *Data file editor* example below).  The `rhandsontable` function also includes the `selectCallback ` parameter, which will fire a callback event when a table cell is selected (see the *Table callback linked to chart* example below).

* [Output only](https://jrowen.shinyapps.io/rhandsontable_output)
```
shiny::runGitHub("rhandsontable", "jrowen", 
                 subdir = "inst/examples/rhandsontable_output")
```
* [Date file editor](https://jrowen.shinyapps.io/rhandsontable_datafile)
```
shiny::runGitHub("rhandsontable", "jrowen", 
                 subdir = "inst/examples/rhandsontable_datafile")
```
* [Calculation input](https://jrowen.shinyapps.io/rhandsontable_portfolio)
```
shiny::runGitHub("rhandsontable", "jrowen", 
                 subdir = "inst/examples/rhandsontable_portfolio")
```
* [Table callback linked to chart](https://jrowen.shinyapps.io/rhandsontable_corr)
```
shiny::runGitHub("rhandsontable", "jrowen", 
                 subdir = "inst/examples/rhandsontable_corr")
```
* [Multiple input tables](https://jrowen.shinyapps.io/rhandsontable_frontier)
```
shiny::runGitHub("rhandsontable", "jrowen", 
                 subdir = "inst/examples/rhandsontable_frontier")
```
* [A shinydashboard app](https://jrowen.shinyapps.io/rhandsontable_dash)
```
shiny::runGitHub("rhandsontable", "jrowen", 
                 subdir="inst/examples/rhandsontable_dash")
```

## Suggestions & Contributions

Please file a issue if you experience any problems with the widget or have feature requests.  Pull requests are also welcome.
